### 2025-08-08 draft.qmd
20250808 - fantasy points by end-of-season position rank and position

```{r}

```

@@

>>>>

```{r}
library("tidyverse")
```

### Load Data {#sec-draftLoadData}

```{r}
#| eval: false
#| include: false

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_seasonal.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_weekly.RData", fsep = ""))
```

```{r}
load(file = "./data/player_stats_seasonal.RData")
load(file = "./data/player_stats_weekly.RData")
```

@@

## Draft Analysis {#sec-draftAnalysis}

### Fantasy Points by (End-of-Season) Position Rank and Position {#sec-expectedPointsByPositionRank}

```{r}
fantasyPointsByPositionRank <- player_stats_seasonal %>% 
  filter(position_group %in% c("QB","RB","WR","TE","K","DL","LB","DB")) %>% 
  group_by(position_group, season) %>% 
  arrange(position_group, season, -fantasyPoints) %>% 
  mutate(positionRank = row_number())
```

A plot of fantasy points by end-of-season position rank and position is depicted in @fig-expectedPointsByPositionRank1.

```{r}
#| label: fig-expectedPointsByPositionRank1
#| fig-cap: "Fantasy Points by End-of-Season Position Rank and Position."
#| fig-alt: "Fantasy Points by End-of-Season Position Rank and Position."
#| code-fold: true

ggplot(
  data = fantasyPointsByPositionRank %>% filter(positionRank <= 200),
  mapping = aes(
    x = positionRank,
    y = fantasyPoints
  )
) + 
  geom_point(
    alpha = 0.3
  ) +
  geom_smooth() +
  coord_cartesian(
    ylim = c(0,NA),
    expand = FALSE) + # don't add space between axes and data
  labs(
    x = "End-of-Season Position Rank",
    y = "Fantasy Points (Season)",
    title = "Fantasy Points (Season) by End-of-Season Position Rank"
  ) +
  theme_bw() +
  facet_wrap(vars(position_group)) # facet by position_group
```

In @fig-expectedPointsByPositionRank2, we overlay the smoothed best-fit line for each position onto the same graphing area for direct comparison.

```{r}
#| label: fig-expectedPointsByPositionRank2
#| fig-cap: "Fantasy Points by End-of-Season Position Rank and Position."
#| fig-alt: "Fantasy Points by End-of-Season Position Rank and Position."
#| code-fold: true

palette_OkabeIto_black <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")

ggplot(
  data = fantasyPointsByPositionRank %>% filter(positionRank <= 200),
  mapping = aes(
    x = positionRank,
    y = fantasyPoints,
    color = position_group
  )
) + 
  geom_smooth(
    linewidth = 1.5,
    se = FALSE) +
  coord_cartesian(
    ylim = c(0,NA), # set limits of y-axis
    expand = FALSE) + # don't add space between axes and data
  scale_color_manual(
    values = palette_OkabeIto_black
  ) +
  labs(
    x = "End-of-Season Position Rank",
    y = "Fantasy Points (Season)",
    title = "Fantasy Points (Season) by End-of-Season Position Rank",
    color = "Position Group"
  ) +
  theme_classic()
```

In @fig-expectedPointsByPositionRank3, we limit the position rank to ranks less than or equal to 30, for easier comparison of positions' comparative dropoff at the top position ranks.

```{r}
#| label: fig-expectedPointsByPositionRank3
#| fig-cap: "Fantasy Points by End-of-Season Position Rank and Position."
#| fig-alt: "Fantasy Points by End-of-Season Position Rank and Position."
#| code-fold: true

ggplot(
  data = fantasyPointsByPositionRank %>% filter(positionRank <= 30),
  mapping = aes(
    x = positionRank,
    y = fantasyPoints,
    color = position_group
  )
) + 
  geom_smooth(
    linewidth = 1.5,
    se = FALSE) +
  coord_cartesian(
    ylim = c(0,425), # set limits of y-axis
    expand = FALSE) + # don't add space between axes and data
  scale_color_manual(
    values = palette_OkabeIto_black
  ) +
  labs(
    x = "End-of-Season Position Rank",
    y = "Fantasy Points (Season)",
    title = "Fantasy Points (Season) by End-of-Season Position Rank",
    subtitle = "Position Rank \u2264 30", # \u2264: <= symbol
    color = "Position Group"
  ) +
  theme_classic()
```

It appears that Running Backs show the fastest dropoff (i.e., steepest negative slope), followed by Tight Ends, Quarterbacks, Wide Receivers, and then defensive positions.
So, identifying the top-scoring Running Back can be of great benefit.

However, the style of gameplay has changed.
@fig-expectedPointsByPositionRank4 depicts the same data since 2020.

```{r}
#| label: fig-expectedPointsByPositionRank4
#| fig-cap: "Fantasy Points by End-of-Season Position Rank and Position."
#| fig-alt: "Fantasy Points by End-of-Season Position Rank and Position."
#| code-fold: true

ggplot(
  data = fantasyPointsByPositionRank %>% filter(positionRank <= 30 & season >= 2020),
  mapping = aes(
    x = positionRank,
    y = fantasyPoints,
    color = position_group
  )
) + 
  geom_smooth(
    linewidth = 1.5,
    se = FALSE) +
  coord_cartesian(
    ylim = c(0,425), # set limits of y-axis
    expand = FALSE) + # don't add space between axes and data
  scale_color_manual(
    values = palette_OkabeIto_black
  ) +
  labs(
    x = "End-of-Season Position Rank",
    y = "Fantasy Points (Season)",
    title = "Fantasy Points (Season) by End-of-Season Position Rank",
    subtitle = "Position Rank \u2264 30, Since 2020", # \u2264: <= symbol
    color = "Position Group"
  ) +
  theme_classic()
```

Although Running Backs appear to show the fastest dropoff, Quarterbacks and Wide Receivers have appeared to overtake them in terms of fantasy points.

@fig-pointsBySeasonByPosition depicts the mean fantasy points of the top 10 players at each position by season.

```{r}
#| label: fig-pointsBySeasonByPosition
#| fig-cap: "Mean Fantasy Points of Top 10 Players at Each Position by Season."
#| fig-alt: "Mean Fantasy Points of Top 10 Players at Each Position by Season."
#| code-fold: true

meanFantasyPointsTop10ByPosition <- fantasyPointsByPositionRank %>% 
  filter(positionRank <= 10) %>% 
  summarise(
    fantasyPoints = mean(fantasyPoints, na.rm = TRUE),
    .groups = "drop_last"
  )

ggplot(
  data = meanFantasyPointsTop10ByPosition,
  mapping = aes(
    x = season,
    y = fantasyPoints,
    color = position_group
  )
) + 
  geom_line(
    linewidth = 1.5) +
  coord_cartesian(
    ylim = c(0,425), # set limits of y-axis
    expand = FALSE) + # don't add space between axes and data
  scale_color_manual(
    values = palette_OkabeIto_black
  ) +
  labs(
    x = "Season",
    y = "Mean Fantasy Points",
    title = "Mean Fantasy Points of Top 10 Players at Each Position by Season",
    color = "Position Group"
  ) +
  theme_classic()
```

The trend suggests that, since 2000, Quarterbacks have increased in fantasy points scored, and Running Backs have decreased in fantasy points scored.
